Hierarchical Bayesian regression modelling predicts biochemical rection energies and facilitates quantitative modelling of cell metabolism

### 2023-10-27 paper/main.tex
Rerender paper

\title[]{Hierarchical Bayesian regression modelling predicts biochemical
rection energies and facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, boxrule=0pt, breakable, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

Generating an artificial dataset required a specification of the true
measurement error; we specified a standard deviation of 0.1 on
logarithmic scale for enzyme and metabolite concentration measurements,
corresponding to approximately 10\% measurement error and a standard
deviation for each reaction measurement approximately 10\% of the
simulated value.

@@

To investigate how the flux of GNMT between conditions {[}X{]} and
{[}Y{]} changes we compared the regulatory contributions in the
posterior distribution. Presented in Figure~\ref{fig-decomposition} is
the regulatory description of GNMT, an irreversible enzyme that is
homotropically activated by its substrate, competitively inhibited by
its product and heterotropically inhibited. The diverse regulation makes
it the ideal test case to elucidate regulatory changes.

Here we found that saturation and allosteric effects were the main
drivers of regulation between the two conditions.
Figure~\ref{fig-decomposition} shows the distributions of the regulatory
components as the log-ratio of each component with {[}X{]} in the
numerator and {[}Y{]} in the denominator, positive values suggest that
this component was increased in experiment {[}X{]} relative to {[}Y{]},
with 0 indicating no difference. Naturally, this presents the ideal
statistical test when evaluating the regulation, with the probability of
regulation given by the relative proportion of samples above and below
the 0 point.

We demonstrate that Maud can accurately reproduce the regulatory effect
between the two conditions by comparing it to the ground truth. As the
true values are well within our probability mass of the posterior we can
say that Maud is able to accurately reproduce the regulatory behaviour
of metabolic networks and also provides a tool for statistical inference
of the regulatory behaviour.

Here we have demonstrated a key feature in Maud, that is, direct
inference on any parameter that arises from the data generating process.
More succinctly, we show the benefit of error propagation that Bayesian
inference provides as researchers can directly use quantiles for
hypothesis testing.

Our tool allows inference directly on the joint posterior distribution
rather than the marginal parameter. The implication is that inference
will be better than that of the marginal distributions due to correlated
parameter values, if the data generating process can be directly
modelled.

>>>>

\title[]{Hierarchical Bayesian regression modelling predicts biochemical
rection energies and facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

Generating an artificial dataset required a specification of the true
measurement error. For enzyme and metabolite concentration measurements
we specified a standard deviation of 0.1 on natural logarithmic scale,
corresponding to approximately 10\% measurement error. For each reaction
measurement a measurement standard deviation of approximately 10\% of
the simulated value.

@@

To demonstrate how Maud's output can be used to yield useful metabolic
insights we used the results of our case study to explain why the flux
of the enzyme \(GNMT\) is higher in condition 1 than in condition 2.
GNMT is an irreversible enzyme that is homotropically activated by its
substrate, competitively inhibited by its product and heterotropically
inhibited. The diverse regulation makes it the ideal test case to
elucidate regulatory changes.

Figure~\ref{fig-decomposition} shows the regulatory description of GNMT,
according to the results our main case study analysis. Each curve shows
the marginal posterior distribution of the ratio of the corresponding
regulatory component in condition 1 compared with condition 2. A
positive value indicates that the component was increased in condition 1
relative to condition 2, with 0 indicating no difference. The
probability, according to our model, of the component acting in each
direction is given by the relative area under the curve on each side of
the 0 point.

Our model correctly inferred that saturation and allosteric effects were
the main drivers of regulation between the two conditions in this case,
with the curves for each component aligning with the ground truth shown
in red.

Importantly, this form of analysis takes into account all modelled
sources of uncertainty, including uncertainty about the true values of
the flux in each condition, and propagates this uncertainty into the
final conclusion. Our result shows that Maud could be used in this
realistic case not only to provide a user with an explanation for an
observed difference in fluxes, but also a reasonable judgement as to the
explanation's robustness.
